---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Savvy - A simple R extension interface using Rust

<!-- badges: start -->

[![](https://img.shields.io/github/actions/workflow/status/yutannihilation/savvy/R-CMD-check.yaml?style=flat-square&logo=github)](https://github.com/yutannihilation/savvy/actions/workflows/R-CMD-check.yaml)
[![](https://img.shields.io/crates/v/savvy.svg?style=flat-square&logo=rust)](https://crates.io/crates/savvy)
[![](https://img.shields.io/docsrs/savvy.svg?style=flat-square&logo=docsdotrs)](https://docs.rs/savvy/latest/)

<!-- badges: end -->

## What the hell is this?? Why do you need another framework when there's extendr?

[extendr](https://extendr.github.io/) is great and ready to use, but it's not perfect in some points (e.g., [error handling][error]) and it's a kind of stuck; extendr is too feature-rich and complex that no one can introduce a big change easily. So, I needed to create a new, simple framework to experiment with. The main goal of savvy is to provide a simpler option other than extendr, not to be a complete alternative to extendr.

[error]: https://github.com/extendr/extendr/issues/278

### Why savvy?

In Japanese, "Rust" is pronounced as `sàbí`(錆). Since the sound is similar, and this framework is intended to be used by R-API-savvy people, I chose this name.

## Example Rust code

With savvy, you can implement Rust function like below to create the corresponding R function `to_upper()`.
As you can see, unlike extendr, this framework is simple in the sense that this does few implicit conversions. For the full details, please read [the crate documentation](https://yutannihilation.github.io/savvy/savvy/index.html).


``` rust
/// Convert to Upper-case
/// 
/// @param x A character vector.
/// @export
#[savvy]
fn to_upper(x: StringSexp) -> savvy::Result<savvy::Sexp> {
    // Use `Owned{type}Sexp` to allocate an R vector for output.
    let mut out = OwnedStringSexp::new(x.len())?;

    for (i, e) in x.iter().enumerate() {
        // To Rust, missing value is an ordinary value. In `&str`'s case, it's just "NA".
        // You have to use `.is_na()` method to distinguish the missing value.
        if e.is_na() {
            // Values need to be set by `set_elt()` one by one.
            out.set_elt(i, <&str>::na())?;
            continue;
        }

        let e_upper = e.to_uppercase();
        out.set_elt(i, e_upper.as_str())?;
    }

    out.into()
}
```

## Getting Started

### Prerequisite

Before starting, install a helper R package for savvy.

``` r
install.packages(
    "savvy",
    repos = c("https://yutannihilation.r-universe.dev", "https://cloud.r-project.org")
)
```

Note that, under the hood, this is just a simple wrapper around `savvy-cli`. So, if you prefer shell, you can directly use the CLI instead, which is available on the [releases](https://github.com/yutannihilation/savvy/releases).

### Create a new R package

First, create a new package. `usethis::create_package()` is convenient for this.

``` r
usethis::create_package("path/to/foo")
```

Then, move to the package directory and generate necessary files like `Makevars` and `Cargo.toml`, as well as generating C and R wrapper code corresponding to the Rust code. `savvy::savvy_init()` does this all (under the hood, this simply runs `savvy-cli init`).

Lastly, run `devtools::document()` to generate `NAMESPACE` and documents.

``` r
savvy::savvy_init()
devtools::document()
```

Now, this package is ready to install!

### Update wrapper files

After modifying or adding some Rust code, you can update the C and R wrapper files by running `savvy::savvy_update()` (under the hood, this simply runs `savvy-cli update`).

``` r
savvy::savvy_update()
devtools::document()
```

## Random thoughts

### Error Handling

This framework uses [tagged pointer](%22https://en.wikipedia.org/wiki/Tagged_pointer%22) to indicate the error. This requires to create some wrappers on C's side to raise an error after checking the tag, but the advantage is that this doesn't need additional allocation and these bit operations should be cheap.

See [my blog post](https://yutani.rbind.io/post/dont-panic-we-can-unwind/) for more details.

### Read-only and writable versions of wrappers

To consider safety first, Rust should not modify the memory allocated by R, and vice versa. So, this framework distinguishes the read-only (or external) data and writable (or owned) data. While this costs some additional memory allocation, this is also good in that we can skip protecting the external data.

This idea comes from [cpp11's `writable`](https://cpp11.r-lib.org/articles/motivations.html#copy-on-write-semantics), but savvy's design is less convenient for educational purposes.

### Do we really want embedded usages?

I've been wondering if it would be simpler if extendr give up supporting embedded usages. Yes, extendr is not only about R packages. It also provides functionalities to embed R in Rust like [this one](https://github.com/yutannihilation/extendr-tide-api-server-example). This is great, but, on the other hand, this means we have to care a lot of things. But, how careful we try to be, concurrency is a tough job.

A lot of tests have failed because of various problems related to concurrency. Every time we encountered such a failure, we place `single_threaded()` here and there. But, could all of them really happen inside an R package? If extendr gives up supporting embedded usages, can our life be simpler a bit?

### Generate bindings by CLI

Extendr embeds the functionality to generate the R bindings and call entries. But, might it be easier to generate them by using an external CLI? I actually need this because I need to generate C code to handle the errors on C's side.

But, static analysis has its pros and cons. While it would be good at parsing a single file `lib.rs`, it's probably hard to understand multiple Rust files correctly. For example, if `misc.rs` defines a function with `#[savvy]` and `lib.rs` imports it, it might be hard to infer how the function is exported. On the other hand, extendr collects the metadata on compile time. So, this sort of problems should never happen.

#### Usage

``` console
Generate C bindings and R bindings for a Rust library

Usage: savvy-cli <COMMAND>

Commands:
  c-header      Generate C header file
  c-impl        Generate C implementation for init.c
  r-impl        Generate R wrapper functions
  makevars      Generate Makevars
  makevars-win  Generate Makevars.win
  gitignore     Generate .gitignore
  update        Update wrappers in an R package
  init          Init savvy-powered Rust crate in an R package
  help          Print this message or the help of the given subcommand(s)

Options:
  -h, --help  Print help
```

## Crates

-   `savvy`: a simple wrapper around R's C API
-   `savvy-bindgen`: a crate for generating C and R bindings from Rust code
-   `savvy-cli`: a CLI to invoke `savvy-bindgen`'s functionalities
-   `savvy-macro`: a crate for `#[savvy]` macro, which is powered by `savvy-bindgen`

## Example functions

A toy example R package can be found in `R-package/` directory.

```{r}
library(savvyExamples)

to_upper(c("a", NA, "A", "座布団一枚"))

times_two_int(c(1L, NA, 100L, 0L, -1L))

times_two_numeric(c(1.1, NA, 0.0, Inf, -Inf))

flip_logical(c(TRUE, FALSE, NA))

x <- Person()

x$set_name("たかし")
x$name()
```

## Rust API

+---------------+--------------------+--------------------+------------------------+
| R type        | Read-only version  | Writable version   | Note                   |
+===============+====================+====================+========================+
| `INTSXP`      | `IntegerSexp`      | `OwnedIntegerSexp` |                        |
+---------------+--------------------+--------------------+------------------------+
| `REALSXP`     | `RealSexp`         | `OwnedRealSexp`    |                        |
+---------------+--------------------+--------------------+------------------------+
| `LGLSXP`      | `LogicalSexp`      | `OwnedLogicalSexp` | -   cannot handle `NA` |
+---------------+--------------------+--------------------+------------------------+
| `STRSXP`      | `StringSexp`       | `OwnedStringSexp`  |                        |
+---------------+--------------------+--------------------+------------------------+
| `VECSXP`      | `ListSexp`         | `OwnedListSexp`    |                        |
+---------------+--------------------+--------------------+------------------------+