---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# unextendr

<!-- badges: start -->
<!-- badges: end -->

## What's this?

This is... nothing. I'm just re-inventing the wheel only to have slightly better
understanding about what [extendr](https://extendr.github.io/) does.

## Random thoughts

### Error Handling

This framework uses [tagged pointer](https://en.wikipedia.org/wiki/Tagged_pointer) to indicate the error.
This requires some wrappers, but the advantage is that we don't need to pass additional data
and these bit operations should be cheap.

But, as (hopefully) [the stabilization of "C-unwind ABI"](https://github.com/rust-lang/rust/issues/74990#issuecomment-1363473645)
is not very far, I'm not sure if this is worth considering to implement in extendr.

See [extendr/extendr#278](https://github.com/extendr/extendr/issues/278) for more discussion.

### Protection

I implemented the doubly-linked list method, which [cpp11 uses](https://cpp11.r-lib.org/articles/internals.html#protection).
Now I'm not sure if this fits extendr; probably its current implementation aims for
parallel processing, so it needs a hashmap to prevent collisions. But, I think
it's not a good idea to use R's C API concurrently anyway, so this should be probably
enough.

### Do we really want embedded usages?

Regarding the concurrency, I'm wondering if it would be simpler if extendr give
up supporting embedded usages. Yes, extendr is not only about R packages. It also
provides functionalities to embed R in Rust like [this one](https://github.com/yutannihilation/extendr-tide-api-server-example).
This is great, but, on the other hand, this means we have to care a lot of things.
But, how careful we try to be, concurrency is a tough job.

A lot of tests have failed because of various problems related to concurrency.
Every time we encountered such a failure, we place `single_threaded()` here and
there. But, could all of them really happen inside an R package? If extendr gives
up supporting embedded usages, can our life be simpler a bit?


## Functions

```{r}
library(unextendr)

to_upper("a")
times_two_int(1L)
times_two_numeric(1.1)
```


## References

* https://notchained.hatenablog.com/entry/2023/01/29/163013 (日本語)
